Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-04-15 doc/thesis/intro.tex
Mostly SMC background

An alternative approach has been the analysis of other networks, which can be
estimated with phylogenetic methods from viral sequence data. Some work focuses
on the \defn{phylogenetic network}, in which two nodes are connected if the
genetic distance between their viral sequences is below some threshold.
Primarily, this work has focused on the detection of \defn{phylogenetic
clusters}, which are groups of individuals whose viral sequences are
significantly more similar to each other's than to the general population's.
The phylogenetic network is informative about ``hotspots'' of transmission and
can be used to identify demographic groups to whom targeted interventions are
likely to have the greatest effect~\autocite{poon2014impact}.  However, this
network may show little to no agreement with a contact data obtained through
epidemiological methods~\autocite{yirrell1998molecular, resik2007limitations,
robinson2013dynamics}, and therefore may be a poor proxy for the contact
network. Other studies~\autocite{brown2011transmission} have investigated the
\defn{transmission network}, which is the subgraph of the contact network
consisting of infected nodes and the edges which led to their
infections~\autocite{welch2011statistical} (\cref{fig:contactnet}, left). It is
possible to estimate the transmission network phylogenetically, although the
methods required for doing so are more sophisticated than for estimating the
phylogenetic network~\autocite{brown2011transmission}. These studies again
mostly focusing on clustering, and also on degree distributions.

@@

The \gls{SIS} algorithm described above aims to sample from a high-dimensional
distribution $\pi(x)$, by sequentially sampling from $d$ distributions of lower
but increasing dimension. \textcite{del2006sequential} developed an
\defn{\gls{SMC} sampler} with an alternative objective: to sample sequentially
from $d$ distributions $\pi_1, \ldots, \pi_d$, all of the same dimension and
defined on the same space. The $\pi_i$ are assumed to form a related sequence,
such as posterior distributions attained by sequentially considering new
evidence. As with \gls{SIS}, we assume that $\pi_i = \gamma_i / Z_i$, where
$\gamma_i$ is known pointwise and the normalizing constant $Z_i$ is unknown.

Both algorithms follow the same general pattern: the target distribution is
attained by progressing through a sequence of intermediate distributions. For
\gls{SIS}, the intermediate distributions are lower-dimensional marginals of
the target, while for the \gls{SMC} sampler, they are of the same dimension and
constitute a smooth progression from an initial distribution to the target. In
both cases, the neighbouring distributions in the sequence are related to each
other in some way, and we can take advantage of that relationship to create a
sequence of importance distributions alongside the sequence of targets. In
\gls{SIS}, the neighbouring marginals $\pi(\vec{x_i})$ and
$\pi(\vec{x_{i+1}})$ were related by the conditional density $\pi(x_i \mid
\vec{x_{i-1}})$, which we used to inform the importance distribution.

The sequence of targets considered by the \gls{SMC} sampler are assumed to be
closely enough related, that we can shift an importance distribution for
$\pi_i$ into one for $\pi_{i+1}$ by perturbing the particles in a systematic
way.

>>>>

An alternative approach has been the analysis of other networks, which can be
estimated with phylogenetic methods from viral sequence data. Some work focuses
on the \defn{phylogenetic network}, in which two nodes are connected if the
genetic distance between their viral sequences is below some threshold.
Primarily, this work has focused on the detection of \defn{phylogenetic
clusters}, which are groups of individuals whose viral sequences are
significantly more similar to each other's than to the general population's.
The phylogenetic network is informative about ``hotspots'' of transmission and
can be used to identify demographic groups to whom targeted interventions are
likely to have the greatest effect~\autocite{poon2015impact}.  However, this
network may show little to no agreement with a contact data obtained through
epidemiological methods~\autocite{yirrell1998molecular, resik2007limitations,
robinson2013dynamics}, and therefore may be a poor proxy for the contact
network. Other studies~\autocite{brown2011transmission} have investigated the
\defn{transmission network}, which is the subgraph of the contact network
consisting of infected nodes and the edges which led to their
infections~\autocite{welch2011statistical} (\cref{fig:contactnet}, left). It is
possible to estimate the transmission network phylogenetically, although the
methods required for doing so are more sophisticated than for estimating the
phylogenetic network~\autocite{brown2011transmission}. These studies again
mostly focusing on clustering, and also on degree distributions.

@@

The \gls{SIS} algorithm described above aims to sample from a high-dimensional
distribution $\pi(x)$, by sequentially sampling from $d$ distributions of lower
but increasing dimension. \textcite{del2006sequential} developed an
\defn{\gls{SMC} sampler} with an alternative objective: to sample sequentially
from $d$ distributions $\pi_1, \ldots, \pi_d$, all of the same dimension and
defined on the same space. The $\pi_i$ are assumed to form a related sequence,
such as posterior distributions attained by sequentially considering new
evidence. As with \gls{SIS}, we assume that $\pi_i(x) = \gamma_i(x) / Z_i$,
where $\gamma_i$ is known pointwise and the normalizing constant $Z_i$ is
unknown.

Both algorithms involve progression through a sequence of related
distributions. For \gls{SIS}, these distributions are lower-dimensional
marginals of the target distribution, while for the \gls{SMC} sampler, they are
of the same dimension and constitute a smooth progression from an initial to a
final distribution. In both cases, the neighbouring distributions in the
sequence are related to each other in some way, and we can take advantage of
that relationship to create a sequence of importance distributions alongside
the sequence of targets. In \gls{SIS}, the neighbouring marginals
$\pi(\vec{x_i})$ and $\pi(\vec{x_{i+1}})$ were related by the conditional
density $\pi(x_i \mid \vec{x_{i-1}})$, which we used to inform the importance
distribution. In \gls{SMC}, the relationship between subsequent distributions
is less explicit, but it is assumed that they are related closely enough that
an importance distribution for $\pi_i$ can be easily transformed into one for
$\pi_{i+1}$. In particular, the sequence of importance distributions $\eta_i$
is constructed as
\begin{align}
  \label{eq:impint}
  \eta_i(x) = \int \eta_{i-1}(\xi) K_i(\xi, x) \d\xi,
\end{align}
where $K_i$ is any Markov kernel and the integral is over the space on which the
$\pi_i$ are defined. The choice of $K_i$ should be based on the percieved
relationship between $\pi_{i-1}$ and $\pi_i$. \textcite{del2006sequential}
propose the use of a \gls{MCMC} kernel with equilibrium distribution $\pi_i$.
That is,
\[
  K_i(\xi, x) = \max\left(1, \frac{q(x, \xi)\pi_i(\xi)}{q(\xi, x)\pi_i(x)}\right),
\]
where $q(\xi, x)$ is a proposal function such as a Gaussian distribution
centered at $\xi$ (see \cref{subsec:mfit}).

Although this method of building up $\eta$ appears straightforward, the
drawback is that the importance distribution itself becomes intractible. In
particular, evaluating $\eta_d(x)$ involves a $d$-dimensional integral of the
type \cref{eq:impint}. As it is necessary to evaluate $\eta(x)$ pointwise to
perform \gls{IS}, this construction appears to have defeated the purpose of
providing an importance distribution for each $\pi_i$.
\textcite{del2006sequential} overcome this problem with two ``artificial''
objects. First, they propose the existence of \textit{backward} kernels
$L_{i-1}(x_i, x_{i-1})$ which perform the inverse operation as $K_i$, namely
transforming $\eta_i$ back into $\eta_{i-1}$. As we shall see, these backward
kernels do not need to be calculated analytically in terms of $K_i$. Second,
they define a sequence of $i$-dimensional target distributions
\[
  \tilde{\pi}_i(\vec{x_i}) = \frac{\gamma_i(x_i)}{Z_i} \prod_{k=1}^{i-1} L_k(x_{k+1}, x_k)
\]
which admit $\pi_i(\vec{x_i})$ as marginals.

### 2016-04-15 doc/thesis/results.Rnw
Mostly SMC background

\Cref{fig:kpca} shows \gls{kPCA} projections of the simulated trees onto the
first two principal components of the kernel matrix. The figure shows only the
simulations with 500-tip trees and, for all parameters except $I$, 1000
infected nodes. The three \gls{alpha} and \gls{I} values considered are well
separated from each other in feature space (\cref{fig:pakpca}). On the other
hand, the three \gls{N} values overlap significantly, and the three \gls{m}
values are virtually indistinguishable. Similar observations can be made for
other values of \gls{I} and the number of tips 
(\cref{fig:alphakpca,fig:Nkpca,fig:Ikpca,fig:mkpca}). The values of \gls{I} and
\gls{N} separate more clearly with larger numbers of tips, and in the case of
\gls{N} larger epidemic sizes.

\begin{figure}[ht]
  \centering
  \includegraphics{kernel-kpca.pdf}
  \caption[\gls{kPCA} projections of simulated trees under varying \gls{BA}
           parameter values]{
    Each parameter of the \gls{BA} model was individually varied to produce 300
    simulated trees. Kernel matrices were formed from all pairwise kernel
    scores among each set of 300 trees. The trees were projected onto the first
    two principal components of the kernel matrix calculated using \gls{kPCA}.
    All trees had 500 tips. The parameters not being varied were set to
    \gls{alpha} = 1, \gls{I} = 1000, \gls{m} = 2, and \gls{N} = 5000. The tree
    kernel meta-parameters were $\lambda = 0.3$ and $\sigma = 4$.
  }
  \label{fig:pakpca}
\end{figure}

@@

Accuracy of the \gls{kSVM} classifiers varied based on the parameter being
tested (\cref{fig:rsquared}, left). Classifiers based on two other tree
statistics, the \gls{nltt} and Sackin's index, generally exhibited worse
performance than the tree kernel, although the magnitude of the disparity
varied between the parameters (\cref{fig:rsquared}, centre and right).
\Cref{fig:rsquared} shows the cross-validation accuracy of each of the tree
classifiers. The results were largely robust to variations in the tree kernel
meta-parameters $\lambda$ and $\sigma$, although accuracy varied between
different epidemic and sampling scenarios
(\cref{fig:alphacrossv,fig:mcrossv,fig:Icrossv,fig:Ncrossv}).

@@

We used grid search to obtain \defn{marginal} estimates for each network
parameter while holding all other parameters fixed. We observed that kernel
scores were highest at the values of \gls{alpha} on the grid closest to the
true values, as shown in Figure~\ref{fig:gridkernel}. However, there was a much
stronger spike in kernel scores near the true value for \gls{alpha} = 1.0 and
1.25. This is recapitulated when we look at the accuracy of point estimates
obtained by taking the grid value with the highest median kernel score. As
shown in Figure~\ref{fig:gridest}, while the estimates are generally close to
the true value, they are much closer for \gls{alpha} = 1.25 than for the other
values.

>>>>

\Cref{fig:kpca} shows \gls{kPCA} projections of the simulated trees onto the
first two principal components of the kernel matrix. The figure shows only the
simulations with 500-tip trees and, for all parameters except $I$, 1000
infected nodes. The three \gls{alpha} and \gls{I} values considered are well
separated from each other in feature space. On the other hand, the three
\gls{N} values overlap significantly, and the three \gls{m} values are
virtually indistinguishable. Similar observations can be made for other values
of \gls{I} and the number of tips
(\cref{fig:alphakpca,fig:Nkpca,fig:Ikpca,fig:mkpca}). The values of \gls{I} and
\gls{N} separate more clearly with larger numbers of tips, and in the case of
\gls{N} larger epidemic sizes.

\begin{figure}[ht]
  \centering
  \includegraphics{kernel-kpca.pdf}
  \caption[\gls{kPCA} projections of simulated trees under varying \gls{BA}
           parameter values]{
    Each parameter of the \gls{BA} model was individually varied to produce 300
    simulated trees. Kernel matrices were formed from all pairwise kernel
    scores among each set of 300 trees. The trees were projected onto the first
    two principal components of the kernel matrix calculated using \gls{kPCA}.
    All trees had 500 tips. The parameters not being varied were set to
    \gls{alpha} = 1, \gls{I} = 1000, \gls{m} = 2, and \gls{N} = 5000. The tree
    kernel meta-parameters were $\lambda = 0.3$ and $\sigma = 4$.
  }
  \label{fig:kpca}
\end{figure}

@@

Accuracy of the \gls{kSVM} classifiers varied based on the parameter being
tested (\cref{fig:rsquared}, left). Classifiers based on two other tree
statistics, the \gls{nltt} and Sackin's index, generally exhibited worse
performance than the tree kernel, although the magnitude of the disparity
varied between the parameters (\cref{fig:rsquared}, centre and right). The
results were largely robust to variations in the tree kernel meta-parameters
$\lambda$ and $\sigma$, although accuracy varied between different epidemic and
sampling scenarios
(\cref{fig:alphacrossv,fig:mcrossv,fig:Icrossv,fig:Ncrossv}).

@@

The accuracy of grid search estimates largely paralleled that of the \gls{kSVM}
classifiers.

We used grid search to obtain marginal estimates for each network parameter
while holding all other parameters fixed. We observed that kernel scores were
highest at the values of \gls{alpha} on the grid closest to the true values, as
shown in Figure~\ref{fig:gridkernel}. However, there was a much stronger spike
in kernel scores near the true value for \gls{alpha} = 1.0 and 1.25. This is
recapitulated when we look at the accuracy of point estimates obtained by
taking the grid value with the highest median kernel score. As shown in
Figure~\ref{fig:gridest}, while the estimates are generally close to the true
value, they are much closer for \gls{alpha} = 1.25 than for the other values.
